---
title: "Student Habits vs Academic Performance"
subtitle: "Data Programming 2025 - Project"
date: "`r format(Sys.time(), '%e %B %Y')`"
author: 
  - name: "Zhivko Stoimchev"
    email: "89221056@student.upr.si"
    affiliations:
    - name: "FAMNIT, University of Primorska"
      city: "Koper"
      state: "Slovenia"
      postal-code: "6000"
format:
  html:
    toc: true
    toc_float: true
    toc-location: left
    embed-resources: true
    theme:
      light: darkly
execute:
  warning: false
  message: false
bibliography: references.bib
---

# Introduction

Students develop all kinds of habits during their school years. Some pull all-nighters before exams, while others stick to regular study schedules. Some prioritize eight hours of sleep, and others survive on much less. Screen time varies too - from minimal use to constant scrolling. These daily choices likely impact academic performance, but which ones matter most?

## The problem

While we often hear advice about good study habits, there's surprisingly little clear evidence about which routines actually lead to better grades. Professors recommend certain approaches, students develop their own methods, and parents suggest what worked in their day. But with so many factors involved - study time, sleep, technology use, and more - it's hard to know what really makes a difference in academic success.

## Objectives

This project aims to:

-   Examine how different study approaches affect grades
-   Measure the relationship between sleep patterns and academic performance
-   Analyze whether screen time helps or hurts school results
-   Provide visual evidence of what habits correlate with success

# Libraries

```{r}
library(tidyverse)
library(skimr)
library(knitr)
library(psych)
library(janitor)
library(ggplot2)
```

# Data presentation

The analysis uses data from 1,000 students to identify which daily routines tend to accompany better grades. This dataset was obtained from Kaggle(@dataset_cit). It does not include any missing values, and all columns are complete. The dataset contains information about students' study habits, social media usage, sleep patterns, and academic performance.

## Data columns

The dataset consists of 1,000 rows and 16 columns. Each row represents a student, and the columns contain information about their study habits and academic performance. The columns are as follows:

| Name | Description | Type |
|---------------------|-----------------------------------|----------------|
| student_id | Unique identifier for each student | Character |
| age | Student's age in years | Numeric |
| gender | Student's gender (Male/Female/Other) | Factor |
| study_hours_per_day | Daily hours spent studying | Numeric |
| social_media_hours | Daily hours spent on social media | Numeric |
| netflix_hours | Daily hours spent watching Netflix | Numeric |
| part_time_job | Whether student has a job (Yes/No) | Logical |
| attendance_percentage | Class attendance rate (0-100%) | Numeric |
| sleep_hours | Average nightly sleep duration | Numeric |
| diet_quality | Self-reported diet quality (Poor/Fair/Good) | Factor |
| exercise_frequency | Weekly exercise sessions | Numeric |
| parental_education_level | Highest parental education | Factor |
| internet_quality | Quality of internet access | Factor |
| mental_health_rating | Self-rated mental health (1-10 scale) | Numeric |
| extracurricular_participation | Whether student does extracurriculars | Logical |
| exam_score | Final exam score (0-100 scale) | Numeric |

## Presentation of the Data

```{r}
data <- read_csv("student_habits_performance.csv")
head(data, 5) %>% kable()
skim(data)
```

Among the 1,000 columns, we can see that there are no missing values.

```{r}
data %>%
  clean_names() %>%
  describe() %>%
  kable()
```

# Transformation of data

Through grouping and normalization, student patterns will be uncovered, revealing how study habits, sleep, and screen time relate to academic performance.

## Categorizing students based on study intensity

```{r}
data_with_study_levels <- data %>%
  mutate(study_intensity = case_when(
    study_hours_per_day < 2 ~ "Light",
    study_hours_per_day >= 2 & study_hours_per_day < 4 ~ "Moderate",
    study_hours_per_day >= 4 ~ "Intense"
  ))

data_with_study_levels %>%
  group_by(study_intensity) %>%
  slice_sample(n = 2) %>%
  select(student_id, study_hours_per_day, study_intensity, exam_score) %>%
  arrange(study_intensity, study_hours_per_day) %>%
  kable(caption = "Representative Students from Each Study Intensity Level")
```

This categorizes students into three study intensity levels based on their daily study hours:

-   **Light**: \< 2 hours
-   **Moderate**: 2 to 4 hours
-   **Intense**: 4 or more hours

We can see that students with more intense study habits tend to perform better academically.

## screen time

```{r}
# Add this in your "Transformation of data" section
# Define a function to categorize screen time usage
categorize_screen_time <- function(social_media, netflix) {
  case_when(
    (social_media + netflix) > 6 ~ "Very High",
    (social_media + netflix) > 4 ~ "High",
    (social_media + netflix) > 2 ~ "Moderate",
    TRUE ~ "Low"
  )
}

# Apply the function to create a new column
data <- data %>%
  mutate(screen_time_level = categorize_screen_time(social_media_hours, netflix_hours))

# Show results
data %>%
  count(screen_time_level) %>%
  kable(caption = "Students by Screen Time Level")
```

## Gender Based Comparison

```{r}
data %>%
  group_by(gender) %>%
  summarize(avg_exam_score = mean(exam_score, na.rm = TRUE),
            avg_study_hours = mean(study_hours_per_day, na.rm = TRUE),
            avg_sleep_hours = mean(sleep_hours, na.rm = TRUE)) %>%
  kable(caption = "Correlation between genders and exam scores")
```

From the results above, we can see that students who identify as "Other" have the highest average exam score (70.65) and the most sleep (6.80 hours), while females study slightly more (3.58 hours) and score higher than males, whose averages are marginally lower in all categories.

## Average Performance per Performance Level

```{r}
data %>%
  mutate(performance_group = case_when(
    exam_score < 50 ~ "Fail",
    exam_score >= 50 & exam_score < 60 ~ "Sufficient",
    exam_score >= 60 & exam_score < 70 ~ "Satisfactory",
    exam_score >= 70 & exam_score < 80 ~ "Good",
    exam_score >= 80 & exam_score < 90 ~ "Very Good",
    exam_score >= 90 & exam_score < 100 ~ "Excellent",
    exam_score == 100 ~ "Outstanding"
  )) -> data
```

Added `performance_group` column to categorize students based on their exam scores. The categories are defined as follows:

-   **Fail**: exam score \< 50
-   **Sufficient**: 50 \<= exam score \< 60
-   **Satisfactory**: 60 \<= exam score \< 70
-   **Good**: 70 \<= exam score \< 80
-   **Very Good**: 80 \<= exam score \< 90
-   **Excellent**: 90 \<= exam score \< 100
-   **Outstanding**: exam score = 100

### Performance Groups Summary

```{r}
data %>%
  group_by(performance_group) %>%
  summarize(n_students = n(),
            avg_exam_score = mean(exam_score, na.rm = TRUE),
            avg_study_hours = mean(study_hours_per_day, na.rm = TRUE),
            avg_attendance_hours = mean(attendance_percentage, na.rm = TRUE),
            avg_sleep_hours = mean(sleep_hours, na.rm = TRUE),
            avg_social_media_hours = mean(social_media_hours, na.rm = TRUE),
            avg_netflix_hours = mean(netflix_hours, na.rm = TRUE)) %>%
  arrange(desc(avg_exam_score)) -> data_stats

data_stats %>% 
    kable()
```

This transformation categorizes students by exam performance and calculates average metrics for each group. The results reveal clear patterns: *higher-performing students study more and use less screen time*, while sleep and attendance show minimal variation across groups.

### The most dominant group

```{r}
data_stats %>%
  mutate(n_percentage = n_students / sum(n_students) * 100) %>%
  select(performance_group, avg_exam_score, avg_study_hours, n_students, n_percentage) %>%
  arrange(desc(n_students)) %>%
  kable(caption = "Performance Groups Overview")
```

The most common performance group is "Good" with 234 students, balancing study hours (3.80) and maintaining solid exam scores (74.40).

### Study hours, attendance rate and exam scores

```{r}
data_stats %>%
  mutate(
    norm_study_hours = scale(avg_study_hours),
    norm_attendance = scale(avg_attendance_hours),
    importance = ifelse(norm_study_hours > norm_attendance, "Study Matters", "Attendance Matters")
  ) %>%
  select(performance_group, avg_exam_score, norm_study_hours, norm_attendance, importance) %>%
  arrange(desc(avg_exam_score)) %>%
  kable(caption = "Correlation between Study Hours, Attendance and Exam Scores")
```

Students who study more tend to perform better. The "Outstanding" group has the highest average study hours (normalized) and the best exam scores, while the "Fail" group has the lowest study hours (normalized) and exam scores. Attendance percentages remain stable across all performance groups, indicating that while attendance is important, it does not directly correlate with higher exam scores **(study time matters more).**

### Sleep hours among all the groups

```{r}
data_stats %>%
  select(performance_group, avg_sleep_hours) %>%
  arrange(desc(avg_sleep_hours)) %>%
  kable(caption = "Correlation between Sleep Hours and Performance Groups")
```

Across all performance groups, sleep hours remain consistent (around 6-6.8 hours), suggesting that sleep does not directly determine academic success.

## Identifying High Screen-Time Students

```{r}
data %>%
  summarize(
    overall_avg_exam_score = mean(exam_score, na.rm = TRUE),
    high_screen_time_avg_exam_score = mean(exam_score[social_media_hours > 5 | netflix_hours > 5], na.rm = TRUE)
  ) %>%
  kable()
```

People who spend more than 5 hours on social media or Netflix have an average exam score of 65.86, which is below the overall average of 69.60. This suggests that high screen time may negatively impact academic performance.

# Exploratory data analysis

## Distribution of Students by Performance Group (Bar Plot)

```{r}
data %>%
  count(performance_group) %>%
  ggplot(aes(x = performance_group, y = n, fill = performance_group)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Distribution of Students by Performance Group", x = "Performance Group", y = "Number of Students")
```

The bar chart displays distribution of students across different performance categories. The highest number of students fall in the group "Good", followed by "Satisfactory". The lowest number appears in the "Outstanding" and "Excellent" groups, indicating a lot of people struggle to achieve top scores. Overall, the trend is that an average student has average to above-average results.

## Study Hours vs Exam Score (Scatter Plot)

```{r}
ggplot(data, aes(x = study_hours_per_day, y = exam_score)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", col = "red", se = FALSE) +
  labs(title = "Study Hours vs Exam Score", x = "Study Hours Per Day", y = "Exam Score")
```

The scatter plot shows a positive correlation between study hours and exam scores. The horizontal axis represents study hours, ranging from 0 to 10 hours per day, while the vertical axis shows exam scores, ranging from 0 to 100. Each blue dot represents a student's performance, indicating how their study time relates to their exam results. The trend by the red line is showing a positive correlation between study hours and exam scores. As study time increases, the exam score tends to improve.

## Sleep Hours by Performance Group (Box Plot)

```{r}
ggplot(data, aes(x = performance_group, y = sleep_hours, fill = performance_group)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Sleep Hours by Performance Group", x = "Performance Group", y = "Sleep Hours")
```

The box plot illustrates the distribution of sleep hours across different performance groups. The horizontal axis represents the performance groups, while the vertical axis shows sleep hours, ranging from 0 to 10. Each box represents the interquartile range (IQR) of sleep hours for each group, with the line inside the box indicating the median. The "Outstanding" group has the highest median sleep hours, while the "Fail" group has the lowest. The plot suggests that students who perform better tend to get more sleep, although all groups average around 6-7 hours.

## Exam Score Distribution by Study Intensity (density plot)

```{r}
ggplot(data_with_study_levels, aes(x = exam_score, fill = study_intensity)) +
  geom_density(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Exam Score Distribution by Study Intensity")
```

This density plot compares exam score distributions across study intensity levels. The horizontal axis represents exam scores, while the vertical axis shows the density of students achieving those scores.

- **Intense studiers** show the highest density of scores above 80, and are right-skewed toward top grades.
- **Moderate studiers** have a more balanced distribution, with peaks around 60-75.
- **Light studiers** have the lowest density of high scores, indicating many struggle to achieve passing grades.

# Conclusion

Up until now, the analysis reveals three key insights about student performance, namely:

1.  **Study time matters most**: Students scoring higher results tend to study more, although attendance rates remain consistent across performance groups.
2.  **Screen time hurts**: Students with high social media or Netflix usage (over 5 hours daily) score lower on exams, meaning entertainment distractions have negative impact on performance.
3.  **Sleep is consistent**: All groups average 6-7 hours, suggesting it's a baseline requirement.

## Visualizing distributions

```{r}
data %>%
  select(exam_score, study_hours_per_day, social_media_hours, 
         netflix_hours, sleep_hours, attendance_percentage) %>%
  pivot_longer(-exam_score, names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value, y = exam_score)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  facet_wrap(~metric, scales = "free_x", ncol = 3) +
  labs(title = "Key Factors vs Exam Scores",
       x = "Value (hours/percentage)",
       y = "Exam Score") +
  theme_minimal()
```

It is worth noticing that the distributions of attendance and sleep hours are almost identical, indicating that students who attend classes regularly also tend to maintain a consistent sleep schedule. This suggests that good attendance may be linked to better overall habits, including sleep

- The analysis shows that study time has the strongest correlation with exam scores, while screen time negatively impacts performance. Sleep duration remains consistent across groups, suggesting it is a baseline requirement for all students.
- Attendance percentage has a minimal impact on exam scores, indicating that while it is important, it does not directly correlate with higher academic performance.
- Netflix hours have a weak negative correlation with exam scores, while Social media hours show a slightly stronger negative trend. 

In summary, the analysis highlights the importance of study habits and their impact on academic performance. Students who dedicate more time to studying tend to achieve better exam scores, while excessive screen time can hinder their success. Sleep duration remains relatively stable across performance groups, suggesting it is a necessary foundation for academic achievement.

